PanGraph: scalable bacterial pan-genome graph construction \\ Supplementary Information

### 2022-09-06 supplementary.tex
feat: continue supplementary

\documentclass[aps,rmp,reprint,superscriptaddress,notitlepage,10pt,onecolumn]{revtex4-1}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath,amsthm,amsfonts,amssymb,amscd}
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{enumerate}
\usepackage[final]{hyperref}

@@

To test the accuracy of \textit{PanGraph} in reconstructing the pangenome graph of a set of isolates we generated artificial data using the procedure described above, obtaining both a set of genome sequences and the underlying real pangenome graph. This was done using standard value of the parameters, but varying the rate of HGT in the interval $h=[0.01,1]$ and the mutation rate $\mu=[0,0.01]$. For each $(h,\mu)$ pair 25 different set of data were generated.\\
On each set of data we executed \textit{PanGraph} with values for the energy parameters $\alpha=0$ and $\beta=0$, and with three different alignment kernels:
\begin{itemize}
    \item \textit{minimap2} with option \textit{asm10}
    \item \textit{minimap2} with option \textit{asm20}
    \item \textit{mmseqs2}
\end{itemize}
Thus obtaining three different pangenome graphs per dataset.\\
To link the mutation rate parameter $\mu$ with the average pairwise SNPs distance $\avg{d}$ between homologous segments in the population we evaluated the average pairwise distance for every pancontig in the pangenome graph comprising more than a single path. For each graph we then perform the weighted average of the divergences, using as weight the length of the panconting. Finally, for every pair of parameters $(\mu,h)$ we average these numbers over the 25 different simulations. Results are displayed in Fig.~\ref{fig:snps-suppl}.\\
While the rate of HGT does not influence the average pairwise divergence of pancontigs, the choice of alignment kernel controls their maximum divergence. \textit{Minimap2} is able to correctly merge pangenome graphs with up to $\avg{d} \sim 7\%$ with the \textit{asm20} option, while \textit{mmseqs2} reaches $\avg{d} \sim 10\%$ at the expense of higher time requirements.\\
By performing a linear fit on the datapoints with $h \leq 0.002$ we are able to recover the conversion factor between the HGT rate and average pairwise divergence of pancontigs in our simulations: $\avg{d} \sim 20.8 \, h$.

@@

Once this link has been established we group simulations by the value of $h$ and check for accuracy comparing the reconstructed pangenome graphs with the ground truth provided by our simulation. In particular for each isolate we consider the breakpoints between different pancontigs that tile the genome, and measure the displacement between the real position of these breakpoints and the position reconstructed by pangraph.\\
To evaluate this displacement it is first necessary to link breakpoints on the real and reconstructed pangenome graph. We formulate this problem as an assignment problem, and solve it numerically using the Hungarian method \cite{kuhn1955hungarian}. For each isolate we call $b_i$ for $i=1,\ldots,I$ the panconting breakpoints on the real pangenome graph, and $b_j$ for $j=1,\ldots,J$ the ones on the reconstructed pangenome graph.\footnote{To avoid artifacts generated by the default threshold distance $d=100$ bp of pangraph, if on any of the two pangenome graphs a pair of breakpoints sits at a distance smaller than this threshold, we remove one of them from the set of breakpoints.} We define a cost matrix $D_{ij} = d(b_i,b_j)$ whose elements are distances on the genome of pairs of breakpoints from the two graphs. We then numerically solve the assignment problem on this matrix, and obtain a set of $K = \min\{ I,J \}$ pairings between breakpoints such that the distance between the pairs is minimal and all the breakpoints of the graph with minimum number of breakpoints have been paired. The average distance between breakpoints in these pairs is the average breakpoint distance of the graph.\\
For each simulation we obtain a number of average breakpoint distances equal to the number of isolates in the simulation. In Fig.~\ref{fig:benchmark-accuracy-suppl} we plot the cumulative distribution of these distances, considering only distances smaller than 1 kbp, stratifying simulations according to the average pairwise divergence of pancontigs inferred from the value of $\mu$. Each panel correspond to a different alignment kernel. As the sequence diversity increases we observe a clear transition. For low diversity most of the breakpoints are inferred to be only a few bp away from their real position, while for highly diverged sequences the position of breakpoints is not precisely inferred and can be hundreds of bp away from their real position. For different alignment kernels this transition occurs at different values of the divergence, with \textit{mmseqs2} being accurate at higher divergence values, as can be visualize from Fig.~XX in the main text.

@@

% TODO: also report average core genome size?
% TODO: avoid talking about PanX?
% TODO: add link with accesion numbers and download?

@@

We built pangenome graphs using five different options for the alignment kernel:
\begin{itemize}
    \item kernel \textit{minimap2} with option \textit{asm10} and default values for the energy parameters $\alpha=100$, $\beta=10$.
    \item kernel \textit{minimap2} with option \textit{asm20} and default values for the energy parameters $\alpha=100$, $\beta=10$.
    \item kernel \textit{mmseqs2} and default values for the energy parameters $\alpha=100$, $\beta=10$.
    \item kernel \textit{minimap2} with option \textit{asm10} and null energy parameters $\alpha=0$, $\beta=0$.
    \item kernel \textit{mmseqs2} and null energy parameters $\alpha=0$, $\beta=0$.
\end{itemize}
From the results described in the previous section we expect \textit{mmseqs2} to be able to merge sequences with higher divergence than \textit{minimap2}. Moreover from the definition of the pseudo-energy \textbf{eq.~1} in the main text, the value of the energy parameters $\beta$ defines an upper threshold for sequence divergence $d \sim 1/\beta$. Homologous sequences with divergence higher than this value will result in positive pseudo-energy and will not be merged in a single pancontig. For $\beta=10$ the critical divergence threshold is $d \sim 10\%$. Executing pangraph with the option $\alpha=0$, $\beta=0$ will result in the merging of more diverged sequences, at the cost of a more fragmented pangenome graph (more but shorter pancontigs).

@@

\begin{figure}[htb]
    \includegraphics[width=\textwidth]{figs_suppl/panx_benchmark_suppl.pdf}
    \caption{{\bf Benchmark on real data}.}
    \label{fig:panx-benchmark-suppl}
\end{figure}

>>>>

\documentclass[aps,rmp,reprint,superscriptaddress,notitlepage,10pt,onecolumn]{revtex4-1}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath,amsthm,amsfonts,amssymb,amscd}
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{enumerate}
\usepackage{xr} % reference to other external latex documents
\usepackage[final]{hyperref}

\externaldocument{main}

@@

To test the accuracy of \textit{PanGraph} in reconstructing the pangenome graph of a set of isolates we generated artificial data using the procedure described above, obtaining both a set of genome sequences and the underlying real pangenome graph. This was done using standard value of the parameters, but varying the rate of HGT in the interval $h=[0.01,1]$ and the mutation rate $\mu=[0,0.01]$. For each $(h,\mu)$ pair 25 different sets of data were generated.\\
On each set of data we executed \textit{PanGraph} with values for the energy parameters $\alpha=0$ and $\beta=0$, and with three different alignment kernels:
\begin{itemize}
    \item \textit{minimap2} with option \textit{asm10}
    \item \textit{minimap2} with option \textit{asm20}
    \item \textit{mmseqs2}
\end{itemize}
Thus obtaining three different pangenome graphs per dataset.\\
To link the mutation rate parameter $\mu$ with the average pairwise SNPs distance $\avg{d}$ between homologous segments in the population we evaluated the average pairwise distance for every pancontig in the pangenome graph comprising more than a single path. For each graph we then perform the weighted average of the divergences, using as weight the length of the panconting. Finally, for every pair of parameters $(\mu,h)$ we average these numbers over the 25 different simulations. Results are displayed in Fig.~\ref{fig:snps-suppl}.\\
While the rate of HGT does not influence the average pairwise divergence of pancontigs, the choice of alignment kernel controls their maximum divergence. \textit{Minimap2} is able to correctly merge pangenome graphs with up to $\avg{d} \sim 7\%$ with the \textit{asm20} option, while \textit{mmseqs2} reaches $\avg{d} \sim 10\%$ at the expense of higher time requirements.\\
By performing a linear fit on the datapoints with $h \leq 0.002$ we are able to recover the conversion factor between the HGT rate and average pairwise divergence of pancontigs in our simulations: $\avg{d} \sim 20.8 \, h$.

@@

Once this link has been established we group simulations by the value of $h$ and check for accuracy comparing the reconstructed pangenome graphs with the ground truth provided by our simulation. In particular for each isolate we consider the breakpoints between different pancontigs that tile the genome, and measure the displacement between the real position of these breakpoints and the position reconstructed by pangraph.\\
To evaluate this displacement it is first necessary to link breakpoints on the real and reconstructed pangenome graph. We formulate this problem as an assignment problem, and solve it numerically using the Hungarian method \cite{kuhn1955hungarian}. For each isolate we call $b_i$ for $i=1,\ldots,I$ the panconting breakpoints on the real pangenome graph, and $b_j$ for $j=1,\ldots,J$ the ones on the reconstructed pangenome graph.\footnote{To avoid artifacts generated by the default threshold distance $d=100$ bp of pangraph, if on any of the two pangenome graphs a pair of breakpoints sits at a distance smaller than this threshold, we remove one of them from the set of breakpoints.} We define a cost matrix $D_{ij} = d(b_i,b_j)$ whose elements are distances on the genome of pairs of breakpoints from the two graphs. We then numerically solve the assignment problem on this matrix, and obtain a set of $K = \min\{ I,J \}$ pairings between breakpoints such that the distance between the pairs is minimal and all the breakpoints of the graph with minimum number of breakpoints have been paired. The average distance between breakpoints in these pairs is the average breakpoint distance of the graph.\\
For each simulation we obtain a number of average breakpoint distances equal to the number of isolates in the simulation. In Fig.~\ref{fig:benchmark-accuracy-suppl} we plot the cumulative distribution of these distances, considering only distances smaller than 1 kbp, stratifying simulations according to the average pairwise divergence of pancontigs inferred from the value of $\mu$. Each panel correspond to a different alignment kernel. As the sequence diversity increases we observe a clear transition. For low diversity most of the breakpoints are inferred to be only a few bp away from their real position, while for highly diverged sequences the position of breakpoints is not precisely inferred and can be hundreds of bp away from their real position. For different alignment kernels this transition occurs at different values of the divergence, with \textit{mmseqs2} being accurate at higher divergence values, as can be visualize from Fig.~\ref{fig:toy-accuracy} in the main text.

@@

% TODO: also report average core genome size?
% TODO: evaluate core genome divergence? Or average mash triangle distance?
% TODO: avoid talking about PanX?
% TODO: add link with accesion numbers and download?

@@

We built pangenome graphs using five different options for the alignment kernel:
\begin{itemize}
    \item kernel \textit{minimap2} with option \textit{asm10} and default values for the energy parameters $\alpha=100$, $\beta=10$.
    \item kernel \textit{minimap2} with option \textit{asm20} and default values for the energy parameters $\alpha=100$, $\beta=10$.
    \item kernel \textit{mmseqs2} and default values for the energy parameters $\alpha=100$, $\beta=10$.
    \item kernel \textit{minimap2} with option \textit{asm10} and null energy parameters $\alpha=0$, $\beta=0$.
    \item kernel \textit{mmseqs2} and null energy parameters $\alpha=0$, $\beta=0$.
\end{itemize}
From the results described in the previous section we expect \textit{mmseqs2} to be able to merge sequences with higher divergence than \textit{minimap2}. Moreover from the definition of the pseudo-energy eq.~\ref{eq:pseudo-energy} in the main text, the value of the energy parameters $\beta$ defines an upper threshold for sequence divergence $d \sim 1/\beta$. Homologous sequences with divergence higher than this value will result in positive pseudo-energy and will not be merged in a single pancontig. For $\beta=10$ the critical divergence threshold is $d \sim 10\%$. Executing pangraph with the option $\alpha=0$, $\beta=0$ will result in the merging of more diverged sequences, at the cost of a more fragmented pangenome graph (more but shorter pancontigs).

@@

\begin{figure}[htb]
    \includegraphics[width=\textwidth]{figs_suppl/panx_benchmark_suppl.pdf}
    \caption{{\bf Benchmark on real data}. We run \textit{PanGraph} on datasets consisting of 5 different bacterial species, and using 5 different options for the alignment kernel as described in the text.
        \textbf{A}: number of isolates for each species.
        \textbf{B}: algoritm wall-time.
        \textbf{C}: maximum memory usage.
        \textbf{D}: n. pancontigs in the resulting pangenome graph.
        \textbf{E}: minimum n. of pancontigs that can add up to 50\% of the total pagenome graph length.
        \textbf{F}: threshold length such that if all pancontigs longer than this threshold are added they cover more than 50\% of the full pangenome graph.
        \textbf{G}: total length of the pangenome graph (sum of all pancontig lengths).
        \textbf{H}: fraction of the pangenome graph that is composed of core pancontigs, i.e. pancontigs that occur exactly once per isolate.
        \textbf{I}: ratio between the length of the pangenome graph and the total length of all chromosomes included in the graph.
    }
    \label{fig:panx-benchmark-suppl}
\end{figure}

% TODO: explain the sense of L50 and N50
